Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-28 paper/paper.tex
Typo fixes

Because we are working with open terms, we need a few additional syntactic
lemmas about reduction. First, a renaming $\xi$ is a generalization of
weakening when working with de Bruijn indices. It consistently renumbers the
variables that appears in terms. If a renamed term has been reduced, we can
always recover the result of the reduction without the renaming.
%
\scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
used more generally. Maybe we should replace it with
that?}\yl{Anti-renaming is mentioned once in Section 6 when we talk abou how confluence with $\eta$
depends on anti-renaming for Par. I don't think that discussion in
Section 6 is very important so I'm fine with replacing it  }
\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}
Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
reduction cannot take any real reduction steps and can only step into a term
itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
sufficient.  In fact, all the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
and the augmentation of neutral terms to \rref{I-Void, I-Eq, I-Bool} can be
immediately discharged by Lemma~\ref{lemma:parnenf}.

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the valuation $[[rho]] = [[idtm]]$ (defined in Figure~\ref{fig:auxdef}), which injects
  variables as terms. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}

@@

To demonstrate the scale of our proof scripts,
Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
code\footnote{calculated by the \texttt{cloc} tool, available from
  \url{https://github.com/AlDanial/cloc}.} for each file of our development,
including the base consistency proof from Section~\ref{sec:logreldep} and
\ref{sec:logrelproof} and the extension to $\beta$-normalization for open and
closed terms from Section~\ref{sec:extension}.  For comparison, we have also
proven syntactic type soundness through subject
reduction\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
progress\footnote{\dotv{syntactic\_soundness.v}{wt\_progress}}.

@@

% First, this definition requires the use of Coq's impredicatve \texttt{Prop}
% sort. In particular, the function $[[F]]$ in \rref{I-Pi} requires the use of
% impredicativity because $[[F]]$ must later be instantiated into something
% defined in terms of the logical relation itself (e.g. in
% Lemma~\ref{lemma:piintroalt}).
In the Coq mechanized proof, we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. The definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resilient to changes, but also provides useful documentation for
readers who wish to understand the mechanized proof. Although automation
performs extensive search, it does not use lemmas or invert derivations that
are not specified in the \texttt{use} or \texttt{inv} flags.

@@

However, % with a little more effort,
it should be possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}.
\section{Related Work}
\label{sec:relatedwork}

>>>>

Because we are working with open terms, we need a few additional syntactic
lemmas about reduction. First, a renaming $\xi$ is a generalization of
weakening when working with simultaneous substitutions. It consistently maps the
variables that appears in terms to other variables. If a renamed term has been reduced, we can
always recover the result of the reduction without the renaming.
%
\scw{This lemma is only used to prove \texttt{wn\_antirenaming}, which is then
used more generally. Maybe we should replace it with
that?}\yl{Anti-renaming is mentioned once in Section 6 when we talk abou how confluence with $\eta$
depends on anti-renaming for Par. I don't think that discussion in
Section 6 is very important so I'm fine with replacing it  }
\begin{lemma}[Par anti-renaming\footnote{\dotv{normalform.v}{Par\_antirenaming}}]
  \label{lemma:parantirenaming} If $[[a < xi > => b0]]$, then there
exists some $[[b]]$ such that $[[b < xi > = b0]]$ and $[[a => b]]$.
\end{lemma}

We can show that parallel reduction preserves $\beta$-normal and
neutral forms.
\begin{lemma}[Par preserves $\beta$-neutral and normal forms\footnote{\dotv{normalform.v}{nf\_ne\_preservation}}]
  \label{lemma:parnenf}
  If $[[a => b]]$, then
  \begin{itemize}
  \item $[[ne a]]$ implies $[[ne b]]$
  \item $[[nf a]]$ implies $[[nf b]]$
  \end{itemize}
\end{lemma}
Lemma~\ref{lemma:parnenf} could have been strengthened to say that if
$[[ne a]]$ or $[[nf a]]$ and $[[a => b]]$, then $[[a = b]]$. Since
$\ottkw{ne}$ and $\ottkw{nf}$ captures terms free of $\beta$ redexes, parallel
reduction cannot take any real reduction steps and can only step into a term
itself. However, for the purpose of our proof, Lemma~\ref{lemma:parnenf} is
sufficient.

All the properties we have shown in
Section~\ref{sec:logreldep} and \ref{sec:logrelproof} before the fundamental
lemma can be proven in the same order, where the new cases due to \rref{I-Ne}
and the augmentation of neutral terms to \rref{I-Void, I-Eq, I-Bool} can be
immediately discharged by Lemma~\ref{lemma:parnenf}.

@@

The normalization property then follows as a corollary of the
fundamental theorem.
\begin{corollary}[Existence of $\beta$-normal forms\footnote{\dotv{soundnessopen.v}{mltt\_normalizing}}]
  \label{corollary:exbetanf}
  If $[[G |- a : A]]$, then $[[wn a]]$ and $[[wn A]]$.
\end{corollary}
\begin{proof}
  By the fundamental lemma, we know that $[[G |= a : A]]$. That is,
  for all $[[rho |= G]]$, there exists some $[[i]]$ and $[[S]]$ such
  that $[[InterpR i A {rho} S]]$ and $[[a {rho} in S]]$.
  We pick the $[[rho]]$ to be the identity substitution $[[idtm]]$, which injects
  variables as terms. The side condition $[[idtm |=
  G]]$ is satisfied since Lemma~\ref{lemma:adequacy} says neutral terms,
  including variables, semantically inhabit any $[[S0]]$ where
  $[[S0]]$ is the interpretation of some type. With our choice of
  $[[rho]]$, we have $[[A {rho}]] = [[A {idtm}]] = [[A]]$ and $[[a {rho}]] = [[a{idtm}]] = [[a]]$. Then we
  know that $[[InterpR i A S]]$ and $[[a in S]]$ for some $[[i]]$ and
  $[[S]]$. By Lemma~\ref{lemma:adequacy}, we
  conclude that $[[wn a]]$ and $[[wn A]]$ respectively.
\end{proof}

@@

To demonstrate the scale of our proof scripts,
Figure~\ref{fig:linecount} shows the number of non-blank, non-comment lines of
code\footnote{calculated by the \texttt{cloc} tool, available from
  \url{https://github.com/AlDanial/cloc}.} for each file of our development,
including the base consistency proof from Section~\ref{sec:logreldep} and
\ref{sec:logrelproof} and the extension to $\beta$-normalization for open and
closed terms from Section~\ref{sec:extension}.

The development with the $\beta\eta$-normalization proof from Section~\ref{sec:betaeta} comprises 1573
lines of non-blank, non-comment lines of code. We choose not to
include it in the chart, since the slight difference in lemma
dependencies makes the comparison in each category less
informative. However, we note that compared to the
$\beta$-normalization extension, the $\beta\eta$ extension has the
same line count in the definition of the logical relation and the
semantic soundness proof. The difference only appears in files related
to untyped reduction and normal forms.

For comparison, we have also
proven syntactic type soundness through subject
reduction\footnote{\dotv{syntactic\_soundness.v}{subject\_reduction}} and
progress\footnote{\dotv{syntactic\_soundness.v}{wt\_progress}}.

@@

% First, this definition requires the use of Coq's impredicatve \texttt{Prop}
% sort. In particular, the function $[[F]]$ in \rref{I-Pi} requires the use of
% impredicativity because $[[F]]$ must later be instantiated into something
% defined in terms of the logical relation itself (e.g. in
% Lemma~\ref{lemma:piintroalt}).
In the Coq mechanized proof, % we encode $[[PowerSet STm]]$ as the type \texttt{tm -> Prop}, a predicate over lambda terms. 
the definition of $[[Interp I i A S]]$ has type
\texttt{Prop}, where \texttt{I} has type \texttt{nat -> tm -> Prop} and \texttt{S} has type \texttt {tm -> Prop}.

@@

The automation provided by CoqHammer not only gives us a proof that is shorter
and more resilient to changes, but also provides useful documentation for
readers who wish to understand the mechanized proof. Although automation
performs extensive search, we can configure it to not use lemmas or invert derivations that
are not specified in the \texttt{use} or \texttt{inv} flags.

@@

However, % with a little more effort,
it should be possible to recover a
precise algorithm. The first step is to define a deterministic
small-step reduction relation that is normalizing; that is, the
relation can always find a normal form if there exists one. A good
candidate is the leftmost-outermost reduction strategy. Its
normalizing property is standard and can be proven using the
factorization technique discussed in
\citet{takahashi-parallel-reduction, factorization-essentially}. By
composing the existence of normal forms and the fact that the
deterministic relation is normalizing, we can derive the
accessibility of the deterministic reduction relation
(\mintinline{coq}{Acc} in Coq), and use the accessibility proof as
an induction metric to define an executable algorithm\footnote{Coq's
  singleton elimination principle allows \mintinline{coq}{Acc}, a
  \texttt{Prop} data type with a single constructor, to be eliminated
  to construct runtime relevant data.}. We leave the proof of the
factorization property and the extraction of a conversion algorithm as
part of our future work.
\section{Related Work}
\label{sec:relatedwork}
